---
title: "Assignment 2: Online Retail Analytics"
author: "Steve Lambert"
output:
  word_document: default
  html_notebook: default
---

# Introduction

This notebook analyzes the Online Retail dataset using R. The dataset contains transaction records from a UK-based online retailer between December 2010 and December 2011. The analysis includes transaction counts by country, transaction value calculations, country-level revenue summaries, date/time exploration, customer analysis, missing data checks, return rate analysis, and product revenue analysis.

# Load the Online Retail Dataset

```{r setup}
# Set options to make output easier to read
options(scipen = 999)

# Load the Online Retail dataset
# Make sure Online_Retail.csv is saved in the same folder as this R Markdown file.
Online_Retail <- read.csv("Online_Retail.csv", fileEncoding = "latin1")

# Display the first few rows of the dataset
head(Online_Retail)

# Display the structure of the dataset
str(Online_Retail)

# Display the number of rows and columns
dim(Online_Retail)
```

The dataset contains `r nrow(Online_Retail)` rows and `r ncol(Online_Retail)` columns.

# Question 1: Transaction Counts by Country

The first question asks for the breakdown of the number of transactions by country, including cancelled transactions. It also asks for the results in total number and percentage, showing only countries that account for more than 1% of total transactions.

```{r question-1}
# Count transactions by country
country_transaction_counts <- as.data.frame(table(Online_Retail$Country))

# Rename columns
names(country_transaction_counts) <- c("Country", "TransactionCount")

# Calculate transaction percentage
country_transaction_counts$TransactionPercent <- 
  country_transaction_counts$TransactionCount / nrow(Online_Retail) * 100

# Keep only countries accounting for more than 1% of total transactions
country_transaction_counts_over_1 <- 
  country_transaction_counts[country_transaction_counts$TransactionPercent > 1, ]

# Sort by transaction count from highest to lowest
country_transaction_counts_over_1 <- 
  country_transaction_counts_over_1[order(-country_transaction_counts_over_1$TransactionCount), ]

# Round percentage for cleaner display
country_transaction_counts_over_1$TransactionPercent <- 
  round(country_transaction_counts_over_1$TransactionPercent, 2)

# Display the results
knitr::kable(country_transaction_counts_over_1)
```

The countries accounting for more than 1% of total transactions are shown above. The United Kingdom accounts for the largest share of transactions.

# Question 2: Create TransactionValue Variable

The second question asks us to create a new variable called `TransactionValue`, which is the product of the existing `Quantity` and `UnitPrice` variables.

```{r question-2}
# Create TransactionValue as Quantity multiplied by UnitPrice
Online_Retail$TransactionValue <- Online_Retail$Quantity * Online_Retail$UnitPrice

# Display the first few rows to confirm that the new variable was added
head(Online_Retail)
```

The `TransactionValue` variable has been added to the dataframe. It represents the total value of each transaction line.

# Question 3: Transaction Values by Country

The third question asks for the breakdown of total transaction values by country. Only countries with total transaction value exceeding 130,000 British pounds should be shown.

```{r question-3}
# Sum transaction values by country
country_transaction_values <- aggregate(
  TransactionValue ~ Country,
  data = Online_Retail,
  FUN = sum
)

# Keep only countries with total transaction value exceeding 130,000
country_transaction_values_over_130k <- 
  country_transaction_values[country_transaction_values$TransactionValue > 130000, ]

# Sort from highest to lowest transaction value
country_transaction_values_over_130k <- 
  country_transaction_values_over_130k[order(-country_transaction_values_over_130k$TransactionValue), ]

# Round transaction value for cleaner display
country_transaction_values_over_130k$TransactionValue <- 
  round(country_transaction_values_over_130k$TransactionValue, 2)

# Display the results
knitr::kable(country_transaction_values_over_130k)
```

The table above shows the countries with total transaction values exceeding 130,000 British pounds.

# Question 4: Optional Golden Question - Date and Time Analysis

This optional question asks us to convert `InvoiceDate` into date/time components and then analyze transaction activity by day, month, and hour.

```{r question-4-date-conversion}
# Convert InvoiceDate into a POSIXlt object
Temp <- strptime(
  Online_Retail$InvoiceDate,
  format = "%m/%d/%Y %H:%M",
  tz = "GMT"
)

# Check the converted date-time variable
head(Temp)

# Create new date, day of week, hour, and month variables
Online_Retail$New_Invoice_Date <- as.Date(Temp)
Online_Retail$Invoice_Day_Week <- weekdays(Online_Retail$New_Invoice_Date)
Online_Retail$New_Invoice_Hour <- as.numeric(format(Temp, "%H"))
Online_Retail$New_Invoice_Month <- as.numeric(format(Temp, "%m"))

# Confirm the new variables were created
head(Online_Retail[, c("InvoiceDate", "New_Invoice_Date", 
                       "Invoice_Day_Week", "New_Invoice_Hour", 
                       "New_Invoice_Month")])
```

## Question 4a: Percentage of Transactions by Day of Week

```{r question-4a}
# Define weekday order
day_order <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")

# Count transactions by day of week
transactions_by_day <- as.data.frame(
  table(factor(Online_Retail$Invoice_Day_Week, levels = day_order))
)

# Rename columns
names(transactions_by_day) <- c("DayOfWeek", "TransactionCount")

# Calculate percentage of transactions by day
transactions_by_day$TransactionPercent <- 
  transactions_by_day$TransactionCount / sum(transactions_by_day$TransactionCount) * 100

# Round percentage
transactions_by_day$TransactionPercent <- round(transactions_by_day$TransactionPercent, 2)

# Display results
knitr::kable(transactions_by_day)
```

The table above shows the percentage of transactions by number of transactions for each day of the week.

## Question 4b: Percentage of Transaction Value by Day of Week

```{r question-4b}
# Sum transaction value by day of week
transaction_value_by_day <- aggregate(
  TransactionValue ~ Invoice_Day_Week,
  data = Online_Retail,
  FUN = sum
)

# Calculate percentage of total transaction value
transaction_value_by_day$TransactionValuePercent <- 
  transaction_value_by_day$TransactionValue / sum(transaction_value_by_day$TransactionValue) * 100

# Order by day of week
transaction_value_by_day$Invoice_Day_Week <- 
  factor(transaction_value_by_day$Invoice_Day_Week, levels = day_order)

transaction_value_by_day <- 
  transaction_value_by_day[order(transaction_value_by_day$Invoice_Day_Week), ]

# Round values for cleaner display
transaction_value_by_day$TransactionValue <- round(transaction_value_by_day$TransactionValue, 2)
transaction_value_by_day$TransactionValuePercent <- round(transaction_value_by_day$TransactionValuePercent, 2)

# Display results
knitr::kable(transaction_value_by_day)
```

The table above shows the percentage of transaction value by day of the week.

## Question 4c: Percentage of Transaction Value by Month

```{r question-4c}
# Sum transaction value by month
transaction_value_by_month <- aggregate(
  TransactionValue ~ New_Invoice_Month,
  data = Online_Retail,
  FUN = sum
)

# Calculate percentage of total transaction value
transaction_value_by_month$TransactionValuePercent <- 
  transaction_value_by_month$TransactionValue / sum(transaction_value_by_month$TransactionValue) * 100

# Sort by month
transaction_value_by_month <- 
  transaction_value_by_month[order(transaction_value_by_month$New_Invoice_Month), ]

# Round values for cleaner display
transaction_value_by_month$TransactionValue <- round(transaction_value_by_month$TransactionValue, 2)
transaction_value_by_month$TransactionValuePercent <- round(transaction_value_by_month$TransactionValuePercent, 2)

# Display results
knitr::kable(transaction_value_by_month)
```

The table above shows the percentage of transaction value by month of the year.

## Question 4d: Date with Highest Number of Transactions from Australia

```{r question-4d}
# Filter records for Australia
australia_records <- Online_Retail[Online_Retail$Country == "Australia", ]

# Count transactions from Australia by date
australia_transactions_by_date <- as.data.frame(table(australia_records$New_Invoice_Date))

# Rename columns
names(australia_transactions_by_date) <- c("InvoiceDate", "TransactionCount")

# Sort by transaction count
australia_transactions_by_date <- 
  australia_transactions_by_date[order(-australia_transactions_by_date$TransactionCount), ]

# Display top dates
head(australia_transactions_by_date)

# Identify the date with the highest number of transactions
top_australia_date <- australia_transactions_by_date[1, ]

top_australia_date
```

The date with the highest number of transactions from Australia is `r top_australia_date$InvoiceDate`, with `r top_australia_date$TransactionCount` transactions.

## Question 4e: Best Two-Hour Maintenance Window

The company needs to shut down the website for two consecutive hours. The IT team is available from 7:00 to 20:00. To minimize customer impact, this analysis identifies the two-hour window with the lowest transaction count.

```{r question-4e}
# Count transactions by hour
transactions_by_hour <- as.data.frame(
  table(factor(Online_Retail$New_Invoice_Hour, levels = 0:23))
)

# Rename columns
names(transactions_by_hour) <- c("Hour", "TransactionCount")

# Convert hour to numeric
transactions_by_hour$Hour <- as.numeric(as.character(transactions_by_hour$Hour))

# Define possible start hours.
# Since maintenance takes two hours and the IT team is available until 20:00,
# the latest start time is 18:00.
available_start_hours <- 7:18

# Calculate transaction count for each two-hour window
two_hour_windows <- data.frame(
  StartHour = available_start_hours,
  EndHour = available_start_hours + 2,
  TransactionCount = sapply(available_start_hours, function(hour_value) {
    sum(transactions_by_hour$TransactionCount[
      transactions_by_hour$Hour %in% c(hour_value, hour_value + 1)
    ])
  })
)

# Sort from lowest to highest transaction count
two_hour_windows <- two_hour_windows[order(two_hour_windows$TransactionCount), ]

# Display possible maintenance windows
knitr::kable(two_hour_windows)

# Select the best maintenance window
best_maintenance_window <- two_hour_windows[1, ]

best_maintenance_window
```

Based on transaction count, the best two-hour maintenance window would start at `r best_maintenance_window$StartHour`:00 and end at `r best_maintenance_window$EndHour`:00, because that window has the lowest number of transactions during the IT team's available hours.

# Question 5: Histogram of Transaction Values from Germany

The fifth question asks us to plot a histogram of transaction values from Germany using the `hist()` function.

```{r question-5}
# Filter records for Germany
germany_transactions <- Online_Retail[Online_Retail$Country == "Germany", ]

# Plot histogram of transaction values from Germany
hist(
  germany_transactions$TransactionValue,
  breaks = 50,
  main = "Histogram of Transaction Values from Germany",
  xlab = "Transaction Value",
  ylab = "Frequency"
)
```

The histogram shows the distribution of transaction values from Germany. Most transaction values appear to be concentrated at lower values, with some larger transaction values creating a long right tail.

# Question 6: Customer Transaction Count and Customer Value

This question asks which customer had the highest number of transactions and which customer had the highest total transaction value.

```{r question-6}
# Remove records with missing CustomerID
valid_customer_records <- Online_Retail[!is.na(Online_Retail$CustomerID), ]

# Count transactions by customer
customer_transaction_counts <- as.data.frame(table(valid_customer_records$CustomerID))

# Rename columns
names(customer_transaction_counts) <- c("CustomerID", "TransactionCount")

# Sort by transaction count
customer_transaction_counts <- 
  customer_transaction_counts[order(-customer_transaction_counts$TransactionCount), ]

# Display top customers by transaction count
head(customer_transaction_counts)

# Identify customer with highest transaction count
top_customer_by_count <- customer_transaction_counts[1, ]

# Sum transaction value by customer
customer_transaction_values <- aggregate(
  TransactionValue ~ CustomerID,
  data = valid_customer_records,
  FUN = sum
)

# Sort by total transaction value
customer_transaction_values <- 
  customer_transaction_values[order(-customer_transaction_values$TransactionValue), ]

# Round transaction value for cleaner display
customer_transaction_values$TransactionValue <- 
  round(customer_transaction_values$TransactionValue, 2)

# Display top customers by transaction value
head(customer_transaction_values)

# Identify most valuable customer
top_customer_by_value <- customer_transaction_values[1, ]

top_customer_by_count
top_customer_by_value
```

The customer with the highest number of transactions is CustomerID `r top_customer_by_count$CustomerID`, with `r top_customer_by_count$TransactionCount` transactions.

The most valuable customer by total transaction value is CustomerID `r top_customer_by_value$CustomerID`, with total transaction value of `r top_customer_by_value$TransactionValue` British pounds.

# Question 7: Percentage of Missing Values by Variable

This question asks for the percentage of missing values for each variable in the dataset.

```{r question-7}
# Calculate percentage of missing values for each variable
missing_value_percent <- data.frame(
  Variable = names(Online_Retail),
  MissingPercent = colMeans(is.na(Online_Retail)) * 100
)

# Round percentage for cleaner display
missing_value_percent$MissingPercent <- round(missing_value_percent$MissingPercent, 4)

# Display results
knitr::kable(missing_value_percent)
```

The table above shows the percentage of missing values for each variable. The largest missingness is in the `CustomerID` field.

# Question 8: Missing CustomerID Records by Country

This question asks for the number of transactions with missing `CustomerID` records by country.

```{r question-8}
# Filter records where CustomerID is missing
missing_customer_records <- Online_Retail[is.na(Online_Retail$CustomerID), ]

# Count missing CustomerID records by country
missing_customerid_by_country <- as.data.frame(table(missing_customer_records$Country))

# Rename columns
names(missing_customerid_by_country) <- c("Country", "MissingCustomerIDCount")

# Sort from highest to lowest missing count
missing_customerid_by_country <- 
  missing_customerid_by_country[order(-missing_customerid_by_country$MissingCustomerIDCount), ]

# Display results
knitr::kable(missing_customerid_by_country)
```

The table above shows the number of transactions with missing `CustomerID` values by country.

# Question 9: Optional Golden Question - Average Days Between Consecutive Shopping

This optional question asks how often customers come back to the website for their next shopping trip. This analysis uses distinct shopping dates per customer, calculates the number of days between consecutive shopping dates, and then calculates the average.

```{r question-9}
# Create a customer/invoice/date dataset
customer_invoice_dates <- unique(
  valid_customer_records[, c("CustomerID", "InvoiceNo", "New_Invoice_Date")]
)

# Calculate differences in days between consecutive shopping dates for each customer
days_between_shopping <- unlist(
  lapply(
    split(customer_invoice_dates$New_Invoice_Date, customer_invoice_dates$CustomerID),
    function(date_values) {
      sorted_dates <- sort(unique(date_values))
      
      if (length(sorted_dates) > 1) {
        return(as.numeric(diff(sorted_dates)))
      } else {
        return(numeric(0))
      }
    }
  )
)

# Calculate average days between consecutive shopping dates
average_days_between_shopping <- mean(days_between_shopping, na.rm = TRUE)

average_days_between_shopping
```

Using distinct shopping dates by customer, customers come back to the website approximately every `r round(average_days_between_shopping, 2)` days on average.

# Question 10: Return Rate for French Customers

This question asks for the return rate for French customers. The assignment defines return rate as the ratio of cancelled transactions over total transactions, where cancelled transactions are identified as transactions with negative `Quantity`.

```{r question-10}
# Filter records for France
france_records <- Online_Retail[Online_Retail$Country == "France", ]

# Count cancelled transactions from France
french_cancelled_transactions <- sum(france_records$Quantity < 0, na.rm = TRUE)

# Count total transactions from France
french_total_transactions <- nrow(france_records)

# Calculate French return rate
french_return_rate <- 
  french_cancelled_transactions / french_total_transactions * 100

# Create summary table
french_return_summary <- data.frame(
  Country = "France",
  CancelledTransactions = french_cancelled_transactions,
  TotalTransactions = french_total_transactions,
  ReturnRatePercent = round(french_return_rate, 2)
)

# Display result
knitr::kable(french_return_summary)
```

The return rate for French customers is approximately `r round(french_return_rate, 2)`%.

# Question 11: Product with Highest Revenue

This question asks which product generated the highest revenue for the retailer. Revenue is calculated as the total sum of `TransactionValue`.

```{r question-11}
# Sum transaction value by product
product_revenue <- aggregate(
  TransactionValue ~ StockCode + Description,
  data = Online_Retail,
  FUN = sum
)

# Sort from highest to lowest revenue
product_revenue <- product_revenue[order(-product_revenue$TransactionValue), ]

# Round revenue for cleaner display
product_revenue$TransactionValue <- round(product_revenue$TransactionValue, 2)

# Display top products
head(product_revenue)

# Identify the product with the highest revenue
top_product <- product_revenue[1, ]

top_product
```

The product or item with the highest total revenue is `r top_product$Description`, with StockCode `r top_product$StockCode`, generating total transaction value of `r top_product$TransactionValue` British pounds.

# Question 12: Unique Customers in the Dataset

This question asks how many unique customers are represented in the dataset. Missing `CustomerID` values are excluded because they do not represent identifiable customers.

```{r question-12}
# Count unique customers, excluding missing CustomerID values
unique_customer_count <- length(unique(na.omit(Online_Retail$CustomerID)))

unique_customer_count
```

There are `r unique_customer_count` unique identifiable customers represented in the dataset.

# Conclusion

This analysis explored the Online Retail dataset using R. The analysis created a transaction value variable, summarized transaction counts and values by country, examined time-based transaction patterns, plotted transaction values for Germany, identified key customers, calculated missing data percentages, reviewed missing CustomerID records, estimated customer return timing, calculated the return rate for French customers, identified the top revenue-generating product, and counted unique customers.